Explorations to Optimize the Parareal Algorithm \\to Solve ODEs in Given Applications

### 2014-12-11 paper.tex
fixed typos and added docs

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We look to achieve strong scaling in parallelizing numerical ODE methods. We
test both parallelism by result space as well as parallelism by time, with the
Parareal algorithm, to solve various ODEs. We measure performance and compare in
terms of accuracy, speedup and efficiency. We provide some theoretical analysis
for the Parareal algorithm's runtime, speedup, error convergence, and stability.
For the Parareal agorithm, we observe preliminary promising but not convincing
results in our small scale tests up to 32 processors on Harvard's Odyssey
computing cluster. We analyze possible reasons as to why our implementation may
not be ideal and some tradeoffs that are taken in the Parareal implementation.
Further optimizations are proposed and rationalized as next steps including
ports of to C++ using BLAS. Our code is written in python using mpi4py. We
compare tests from various data sources from basic exponential ODEs to larger
grids of ODEs to solve for in applications like diffusion and sound wave
propagation.

@@

\subsection{Numerical Methods and Parallelization}
Solving systems of differential equations can be a computationally expensive
task. The error of most algorithms scales on the stepsize of the discretization
of time. However, stepsize in time is also proportional to the computation
required. It would be nice to allow for strong scaling where a problem can be
solved in a reasonable time for very small time steps.

@@

An alternative paradigm to applying parallelism to numerical methods is the
parallelism by space paradigm, where the result vector space is divided up to
different regions and sent to different responsible processors. In this way,
each processor sees a similar but smaller-scale problem---weak scaling. This
method is expected to outperform the parallel in time paradigm simply because
the task is embarassingly parallel, being able to be naturally divided
into smaller but identical in method problems. There is no serial component in
this algorithm and should be exhibit both strong and weak scaling.

@@

Performance of parallel algorithms are usually measured in terms of 3 different
metrics: speedup, strong scalling efficiency and weak scaling efficiency.
Strong and weak scaling are common measures for the different ways a program can
benefit from increased parallellism and more processors as well as the
limitations that dictate the speedups~\cite{parallelscaling}. In each of the below subsections, $T(n)$ represents the computation of using $n$ processors and the same notation for speedup.

@@

The final principle when discussing parallel performance is the potential existence of an Amdahl speedup limit. Amdahl's Law seeks to describe the maximum parallel speedup in an overall piece of code when the problem is not embarassingly parallel --- not all of the system can be divided up in parallel. The parallel in time methods must always have some serial component due to the time-dependent nature of the ODE's which will fall under the limitations stated by Amdahl's Law. His law expresses both best time given $n$ processors as well as the speedup for $n$ processors as a function of $B$, the fractional amount of the algorithm that cannot be parallelized---due to strict serial constraints:

@@

To delve one step deeper in our analysis, parallel communications are usually expressed by architecture parameters---$\alpha$, $\beta$ and $\gamma$ explained in the below table. The expressions measure the communitation times spent by each processor (on average as not all processors may communicate equally depending on the size of the problem). This helps model the communication overhead.

@@

\subsubsection{Parallel By Space}
The main communication of the parallel in space method is the sendrecv (a round robin style send and receive for the ghost pounts - the corners have 2 pairs of sendrecvs) which contribute to 4 $\alpha$ plus the initial scatter and final gather each with an $\alpha$ term.  The amount comunicated is the answer as well as the original broadcasted space which includes boundary points. Then, as before, there will be the gamma term multiplied into the computation of the component - given again by an abstract function $\xi(P_n)$ which returns the cost in flops of computing on the input block, $P_n$. 
\[
C_{\textnormal{paraspace}} \leq \alpha \cdot 6 + \beta \cdot P/n + \gamma \cdot \xi(P/n)
\]

@@

Because this task is embarassingly parallel and nearly 100\% of the computation code can be parallelized (disregarding any uneven divisions of the space by the number of processors), we achieve the Amdahl's Law optimal speedup bound of (with $B$ of 0):

@@

\subsection{Expectations}
We hoped to show that under the properly tuned parameters (namely the $k$ number of iterations), we would get error for the Parareal method that would approach that of running $g_{\textnormal{fine}}$. There would be some convergence above which increasing $k$ would not help convergence by as much. The expectation was that we would only being to see the best speedups for the Parareal algorithm with large $n$ due to the strong scaling. With limited access to processors, we expected to see speedup comparing the Parareal algorithm to itself but not necessaily to the evaluation of $g_{\textnormal{fine}}$. Finally, we fully expect the parallel in space algorithm to demonstrate good weak scaling, as well as for it to be faster overall since it is a more embarassingly parallel design. As disclaimed before, the main tradeoff for the parallel in space method is the lack of extensiblily across different ODEs applications.

>>>>

% Put your stuff here
%\twocolumn[
%\begin{@twocolumnfalse}
\maketitle
\begin{abstract}
We look to achieve strong scaling in parallelizing numerical ODE methods. We
test both parallelism by result space as well as parallelism by time, with the
Parareal algorithm, to solve various ODEs. We measure performance and compare in
terms of accuracy, speedup and efficiency. We provide some theoretical analysis
for the Parareal algorithm's runtime, speedup, error convergence, and stability.
For the Parareal algorithm, we observe preliminary promising but not convincing
results in our small scale tests up to 32 processors on Harvard's Odyssey
computing cluster. We analyze possible reasons as to why our implementation may
not be ideal and some tradeoffs that are taken in the Parareal implementation.
Further optimizations are proposed and rationalized as next steps including
ports of to C++ using BLAS. Our code is written in Python using mpi4py. We
compare tests from various data sources from basic exponential ODEs to larger
grids of ODEs to solve for in applications like sound wave propagation.

@@

\subsection{Numerical Methods and Parallelization}
Solving systems of differential equations can be a computationally expensive
task. The error of most algorithms scales on the step size of the discretization
of time. However, step size in time is also proportional to the computation
required. It would be nice to allow for strong scaling where a problem can be
solved in a reasonable time for very small time steps.

@@

An alternative paradigm to applying parallelism to numerical methods is the
parallelism by space paradigm, where the result vector space is divided up to
different regions and sent to different responsible processors. In this way,
each processor sees a similar but smaller-scale problem---weak scaling. This
method is expected to outperform the parallel in time paradigm simply because
the task is embarrassingly parallel, being able to be naturally divided
into smaller but identical in method problems. There is no serial component in
this algorithm and should be exhibit both strong and weak scaling.

@@

Performance of parallel algorithms are usually measured in terms of 3 different
metrics: speedup, strong scaling efficiency and weak scaling efficiency.
Strong and weak scaling are common measures for the different ways a program can
benefit from increased parallelism and more processors as well as the
limitations that dictate the speedups~\cite{parallelscaling}. In each of the below subsections, $T(n)$ represents the computation of using $n$ processors and the same notation for speedup.

@@

The final principle when discussing parallel performance is the potential existence of an Amdahl speedup limit. Amdahl's Law seeks to describe the maximum parallel speedup in an overall piece of code when the problem is not embarrassingly parallel---not all of the system can be divided up in parallel. The parallel in time methods must always have some serial component due to the time-dependent nature of the ODE's which will fall under the limitations stated by Amdahl's Law. His law expresses both best time given $n$ processors as well as the speedup for $n$ processors as a function of $B$, the fractional amount of the algorithm that cannot be parallelized---due to strict serial constraints:

@@

To delve one step deeper in our analysis, parallel communications are usually expressed by architecture parameters---$\alpha$, $\beta$ and $\gamma$ explained in the below table. The expressions measure the communication times spent by each processor (on average as not all processors may communicate equally depending on the size of the problem). This helps model the communication overhead.

@@

\subsubsection{Parallel By Space}
The main communication of the parallel in space method is the sendrecv (a round
robin style send and receive for the ghost points - the corners have 2 pairs of
sendrecvs) which contribute to 4 $\alpha$ plus the initial scatter and final
gather each with an $\alpha$ term. The amount communicated is the answer as well
as the original broadcasted space which includes boundary points. Then, as
before, there will be the gamma term multiplied into the computation of the
component---given again by an abstract function $\xi(P_n)$ which returns the
cost in flops of computing on the input block, $P_n$. 
\[
C_{\textnormal{paraspace}} \leq \alpha \cdot 6 + \beta \cdot P/n + \gamma \cdot \xi(P/n)
\]

@@

Because this task is embarrassingly parallel and nearly 100\% of the computation code can be parallelized (disregarding any uneven divisions of the space by the number of processors), we achieve the Amdahl's Law optimal speedup bound of (with $B$ of 0):

@@

\subsection{Expectations}
We hoped to show that under the properly tuned parameters (namely the $k$ number
of iterations), we would get error for the Parareal method that would approach
that of running $g_{\textnormal{fine}}$. There would be some convergence above
which increasing $k$ would not help convergence by as much. The expectation was
that we would only being to see the best speedups for the Parareal algorithm
with large $n$ due to the strong scaling. With limited access to processors, we
expected to see speedup comparing the Parareal algorithm to itself but not
necessarily to the evaluation of $g_{\textnormal{fine}}$. Finally, we fully
expect the parallel in space algorithm to demonstrate good weak scaling, as well
as for it to be faster overall since it is a more embarrassingly parallel design.
As disclaimed before, the main tradeoff for the parallel in space method is the
lack of extensibility across different ODEs applications.
